---
title: "UBER PICKUPS'S DATA SCIENCE PROJECT"
output: html_notebook
---

### First, import the neccesary packages.
```{r}
library(ggplot2)
library(ggthemes)
library(lubridate)
library(dplyr)
library(tidyr)
library(DT)
library(scales)
```
 
### Creating a vector of colors that will be included in our plotting functions.

```{r}
colors = c("#CC1011", "#665555", "#05a399", "#cfcaca", "#f5e840", "#0683c9", "#e075b0")
```

## LOAD THE DATA AND PRE-PROCESSING.

### Read several csv files that contain the data from April 2014 to September 2014 and store them in corresponding data frames.

```{r}
apr_data <- read.csv("uber-raw-data-apr14.csv")
may_data <- read.csv("uber-raw-data-may14.csv")
jun_data <- read.csv("uber-raw-data-jun14.csv")
jul_data <- read.csv("uber-raw-data-jul14.csv")
aug_data <- read.csv("uber-raw-data-aug14.csv")
sep_data <- read.csv("uber-raw-data-sep14.csv")
```

### Bind all previous dataframes per column aggregation in a new dataframe called data_2014.

```{r}
data_2014 <- rbind(apr_data,may_data, jun_data, jul_data, aug_data, sep_data)  
View(data_2014)
```

### This operation is commonly used to ensure that the "Date.Time" column is recognized as a date-time object in R.

```{r}
data_2014$Date.Time <- as.POSIXct(data_2014$Date.Time, format = "%m/%d/%Y %H:%M:%S")
View(data_2014)
```

### Create a new column named "Time" in the data_2014 dataframe, and populating it with the time component extracted from the "Date.Time" column. The format function is used to achieve this.

```{r}
data_2014$Time <- format(as.POSIXct(data_2014$Date.Time, format = "%m/%d/%Y %H:%M:%S"), 
                         format="%H:%M:%S")
View(data_2014)
```

### Transform the data in the same way as the previous code but with the package ymd_hms of lubridate.

```{r}
data_2014$Date.Time <- ymd_hms(data_2014$Date.Time)
View(data_2014)
```

### Create a day as a factor and add it as a new column. Day is a function of lubridate package.

```{r}
data_2014$day <- factor(day(data_2014$Date.Time))
View(data_2014)
```

### The same transformation for the month. label=true indicates the month name as a label instead numerical value.

```{r}
data_2014$month <- factor(month(data_2014$Date.Time, label = TRUE))
View(data_2014)
```

### The same transformation for the year.

```{r}
data_2014$year <- factor(year(data_2014$Date.Time))
```

### The same transformation for days of the week.

```{r}
data_2014$dayofweek <- factor(wday(data_2014$Date.Time, label = TRUE))
View(data_2014)
```

### Creation and transformations for hour, minute and second.


```{r}
data_2014$Hour <- factor(hour(hms(data_2014$Time)))
data_2014$minute <- factor(minute(hms(data_2014$Time)))
data_2014$second <- factor(second(hms(data_2014$Time)))
View(data_2014)
```

## PLOTTING THE TRIPS BY DIFFERENT VARIABLES

### Create a summary dataframe called hour_data. Summarizing the totalcount of observations for each unique hour using the summarize function.

```{r}
hour_data <- data_2014 %>% group_by(Hour) %>% dplyr::summarize(Total = n())  
datatable(hour_data)
```


### Plotting with ggplot. "identity" argument indicates that the "Total" values should be used directly.scale_y_continuous(labels = comma) formats the y-axis labels using commas for better readability. 

```{r}
ggplot(hour_data, aes(Hour, Total)) + 
  geom_bar( stat = "identity", fill = "lightgreen", color = "black")+
  ggtitle("Trips Every Hour") + theme(legend.position = "none") +
  scale_y_continuous(labels = comma)
```


### Create a summary dataframe called month_hour. Summarizing the total count of observations for each unique hour and by months using the summarize function.

```{r}
month_hour <- data_2014 %>% group_by(month, Hour) %>% dplyr::summarize(Total = n())
datatable(month_hour)
```

### Plotting with ggplot. "identity" argument indicates that the "Total" values should be used directly. scale_y_continuous(labels = comma) formats the y-axis labels using commas for better readability. In this case, fill = month indicates that the colors of bars will be filled with the variation of months.

```{r}
ggplot(month_hour, aes(Hour, Total, fill = month)) + 
  geom_bar( stat = "identity") + ggtitle("Trips by Hour and Month") + 
  scale_y_continuous(labels = comma)
```


### Create a summary dataframe called day_group. Summarizing the total count of observations for each unique day using the summarize function.

```{r}
day_group <- data_2014 %>% group_by(day) %>% dplyr::summarize(Total = n()) 
datatable(day_group)
```

### Plotting with ggplot. "identity" argument indicates that the "Total" values  should be used directly. scale_y_continuous(labels = comma) formats the y-axis labels using commas for better readability. In this case, fill = orange in order to fill the bar with that particular color.

```{r}
ggplot(day_group, aes(day, Total)) + 
  geom_bar( stat = "identity", fill = "orange") + ggtitle("Trips Every Day") +
  theme(legend.position = "none") + scale_y_continuous(labels = comma)
```


### Create a summary dataframe called day_month_group. Summarizing the total count of observations for each unique day and months using the summarize function.

```{r}
day_month_group <- data_2014 %>% group_by(month, day) %>% dplyr::summarize(Total = n())
```

### Plotting with ggplot. "identity" argument indicates that the "Total" values should be used directly. scale_y_continuous(labels = comma) formats the y-axis labels using commas for better readability. For this graph we use the color vector previously created and apply them to the bars.

```{r}
ggplot(day_month_group, aes(day, Total, fill = month)) +  geom_bar( stat = "identity") +
  ggtitle("Trips by Day and Month") + scale_y_continuous(labels = comma) +
  scale_fill_manual(values = colors)
```

### Create a summary dataframe called month_group. Summarizing the total count of observations for each unique month using the summarize function.

```{r}
month_group <- data_2014 %>% group_by(month) %>% dplyr::summarize(Total = n()) 
datatable(month_group)
```

### Plotting with ggplot. "identity" argument indicates that the "Total" values should be used directly. scale_y_continuous(labels = comma) formats the y-axis labels using commas for better readability. For this graph we use the color vector previously created and apply them to the bars.

```{r}
ggplot(month_group, aes(month, Total, fill = month)) + geom_bar( stat = "identity") +
  ggtitle("Trips by Month") + theme(legend.position = "none") +
  scale_y_continuous(labels = comma) + scale_fill_manual(values = colors)
```

### Create a summary dataframe called month_weekday. Summarizing the total count of observations for each unique day of the week and month using the summarize function.

```{r}
month_weekday <- data_2014 %>% group_by(month, dayofweek) %>% dplyr::summarize(Total = n())
```

### Plotting with ggplot. "identity" argument indicates that the "Total" values should be used directly. scale_y_continuous(labels = comma) formats the y-axis labels using commas for better readability. For this graph we use the color vector previously created and apply them to the bars and position = "dodge": Dodges the bars for different weekdays side by side.

```{r}
ggplot(month_weekday, aes(month, Total, fill = dayofweek)) + geom_bar( stat = "identity", position = "dodge") +
  ggtitle("Trips by Day and Month") + scale_y_continuous(labels = comma) +
  scale_fill_manual(values = colors)
```


### These plots visualize the distribution of trips based on different factors such as bases, months, and days of the week. The color aesthetics add additional information for the second and third plots.

```{r}
ggplot(data_2014, aes(Base)) +  geom_bar(fill = "grey") +
  scale_y_continuous(labels = comma) + ggtitle("Trips by Bases")

ggplot(data_2014, aes(Base, fill = month)) +  geom_bar(position = "dodge") +
  scale_y_continuous(labels = comma) + ggtitle("Trips by Bases and Month") +
  scale_fill_manual(values = colors)

ggplot(data_2014, aes(Base, fill = dayofweek)) + geom_bar(position = "dodge") +
  scale_y_continuous(labels = comma) + ggtitle("Trips by Bases and DayofWeek") +
  scale_fill_manual(values = colors)
```

## CREATING A HEATMAP VISUALIZATION OF DAY, HOUR AND MONTH

### Create a summary dataframe called day_and_hour. Summarizing the total count of observations for each unique day and hour using the summarize function.

```{r}
day_and_hour <- data_2014 %>% group_by(day, Hour) %>% dplyr::summarize(Total = n())
datatable(day_and_hour)
```


### These heat maps visualize the distribution of total values across different combinations of days, hours, months, and days of the week. The color intensity in each tile represents the total value for the corresponding combination of variables. The tiles are filled with the values of Total.

```{r}
ggplot(day_and_hour, aes(day, Hour, fill = Total)) +
  geom_tile(color = "white") + ggtitle("Heat Map by Hour and Day")

ggplot(day_month_group, aes(day, month, fill = Total)) +
  geom_tile(color = "white") + ggtitle("Heat Map by Month and Day")

ggplot(month_weekday, aes(dayofweek, month, fill = Total)) +
  geom_tile(color = "white") + ggtitle("Heat Map by Month and Day of Week")
```


### Create a summary dataframe called month_base. Summarizing the total count of observations for each base and months using the summarize function.

```{r}
month_base <-  data_2014 %>% group_by(Base, month) %>% dplyr::summarize(Total = n())
```

### Create a summary dataframe called month_base. Summarizing the total count of observations for each base and days of the weeks using the summarize function.

```{r}
day0fweek_bases <-  data_2014 %>% group_by(Base, dayofweek) %>% dplyr::summarize(Total = n()) 
```


### These heat maps visualize the distribution of total values across different combinations of bases, months, and days of the week. The color intensity in each tile represents the total value for the corresponding combination of variables. The white borders between tiles enhance the visibility of individual cells in the heat map.

```{r}
ggplot(month_base, aes(Base, month, fill = Total)) +
  geom_tile(color = "white") + ggtitle("Heat Map by Month and Bases")

ggplot(day0fweek_bases, aes(Base, dayofweek, fill = Total)) +
  geom_tile(color = "white") + ggtitle("Heat Map by Bases and Day of Week")
```

## MAP VISUALIZATION IN NEW YORK


### These variables are commonly used to define a bounding box or geographical extent. In this case, the bounding box is defined by the minimum and maximum latitude and longitude values, creating a rectangular area on a map.

```{r}
min_lat <- 40.5774
max_lat <- 40.9176
min_long <- -74.15
max_long <- -73.7004
```


### The code generates a scatter plot with blue points representing Uber ride locations in New York City during the specified months of 2014. The x and y axes are limited to the geographical extent defined by the minimum and maximum longitude and latitude values. The plot has a map-themed style and a title indicating the time period.

```{r}
ggplot(data_2014, aes(x=Lon, y=Lat)) + geom_point(size=1, color = "blue") +
  scale_x_continuous(limits=c(min_long, max_long)) +
  scale_y_continuous(limits=c(min_lat, max_lat)) +
  theme_map() + ggtitle("NYC MAP BASED ON UBER RIDES DURING 2014 (APR-SEP)")
```


### This code generates a scatter plot with points representing Uber ride locations in New York City during the specified months of 2014. The points are differentiated by color based on the "Base" variable. The x and y axes are limited to the geographical extent defined by the minimum and maximum longitude and latitude values. The plot has a map-themed style and a title indicating the time period and the differentiation by base.

```{r}
ggplot(data_2014, aes(x=Lon, y=Lat, color = Base)) +
  geom_point(size=1) + scale_x_continuous(limits=c(min_long, max_long)) +
  scale_y_continuous(limits=c(min_lat, max_lat)) +
  theme_map() + ggtitle("NYC MAP BASED ON UBER RIDES DURING 2014 (APR-SEP) by BASE")
```
